Tutorial_1b

### 2024-10-16 Tutorial_1b.Rmd
add tut1b

---
title: "Tutorial_4"
author: "Sven Panis"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

@@

Tutorial_4 illustrates how to use functional programming to set up life tables, and plot the discrete-time hazard, survivor, and conditional accuracy functions, for the data of Experiment 1 of Panis and Schmidt (2016) when there are TWO independent variables (here: mask type and prime type).

@@

```{r read-raw-data}
raw <- read_delim(file="Tutorial_1_descriptive_stats/data/DataExp1_6subjects.txt", delim="\t",col_names=T) # 9372 rows
head(raw)
tail(raw)
glimpse(raw)
```

@@

# Apply functional programming to set up life tables, and plot the discrete-time functions (h(t), S(t), and ca(t)).

@@

```{r load-functions}
# function censor creates censored observations and discrete response times (drt), given a user-defined censoring time (timeout) and bin width in ms
censor <- function(df, timeout, bin_width){
  if(!(timeout %% bin_width == 0)){
    return("The censoring time must be a multiple of the bin width!")
  }
  if(timeout < 0 | bin_width < 0){
    return("Both timeout and bin_width must be larger than 0!")
  }
  df %>% mutate(right_censored = 0,
                rtc = ifelse(rt > timeout, timeout, rt) %>% round(digits=2), # censored response times
                right_censored = ifelse(rtc == timeout,1,right_censored), # 1 = right censored observation, 0 = observed rt
                drt = ceiling(rtc/bin_width), # drt = discrete response time
                cens_time = timeout, bin_width = bin_width) # save both user-defined parameters for plotting
}

# function ptb creates a person-trial-bin oriented data set
ptb <- function(df){
  df %>% uncount(weights = drt) %>% # create drt rows per trial
         group_by(trial) %>% 
         mutate(period = 1:n()) %>% # create time bin (or time period) ranks within each trial
         mutate(event = if_else(period == max(period) & right_censored == 0, 1, 0)) %>% # event = 1 indicates response occurrence
         ungroup()
}

# function setup_lt sets up a life table for each level 2 independent variables
setup_lt_2IV <- function(ptb){
  ptb %>% mutate(event = str_c("event", event)) %>%
          group_by(condition1,condition2,period) %>% 
          count(event) %>% 
          ungroup() %>% 
          pivot_wider(names_from = event,
                      values_from = n) %>% 
          mutate(event0 = ifelse(is.na(event0),0,event0), # replace NA with 0
                 event1 = ifelse(is.na(event1),0,event1),
                 risk_set = event0 + event1) %>% # define the risk set
          mutate(hazard = (event1 / risk_set) %>% round(digits = 3)) %>% # calculate hazard estimate
          mutate(se_haz = sqrt((hazard * (1 - hazard)) / risk_set) %>% round(digits = 4)) %>% # standard error for hazard
          group_by(condition1,condition2) %>%
          mutate(survival = (cumprod(1-hazard)) %>% round(digits = 4), # calculate survival estimate
                 term     = (cumsum(hazard / (risk_set * (1 - hazard)))) %>% round(digits = 7), # intermediate calculation
                 se_surv  = (survival * sqrt(term)) %>% round(digits = 5)  ) %>% # Greenwood's (1926) approximation
          ungroup() 
}

# function calc_ca calculates the conditional accuracies 
calc_ca_2IV <- function(df){
  df %>% filter(right_censored==0) %>%
         group_by(condition1,condition2,drt,cens_time,bin_width) %>%
         summarize(ca = mean(acc) %>% round(digits = 2),
                   n = n(),
                   .groups = 'drop') %>%
         ungroup() %>%
         mutate(period = drt,
                se_ca = sqrt((ca * (1-ca)) / n) %>% round(digits = 3)) %>%
         select(-drt)
}

# function join_lt_ca_2IV joins the conditional accuracies to the life tables when there are two independent variables
join_lt_ca_2IV <- function(df1,df2){df1 %>% left_join(df2, join_by(condition1,condition2,period))}

# function extract_median_2IV is used to extract the S(t).50 quantiles when there are two independent variables
extract_median_2IV <- function(df){
  above_pct50 <- df %>% 
      group_by(condition1,condition2) %>%
      filter(survival >= .5) %>% 
      slice(n()) # take last row
  below_pct50 <- df %>% 
      group_by(condition1,condition2) %>%
      filter(survival < .5) %>% 
      slice(1) # take first row
  # pull period above
  period_above <- pull(above_pct50, period)
  # pull survivor function values
  survival_above <- pull(above_pct50, survival)
  survival_below <- pull(below_pct50, survival)
  # estimate median by interpolation
  median_period <- period_above+((survival_above-.5)/(survival_above-survival_below))*((period_above+1)-period_above)
}

# function plot_eha_2IV plots the hazard, survivor, and ca(t) functions, when there are two independent variables
plot_eha_2IV <- function(df,subj,haz_yaxis){ 
  library(patchwork)
  cutoff <- df %>% pull(cens_time) %>% max(na.rm=T)
  binsize <- df %>% pull(bin_width) %>% max(na.rm=T)
  median_period <- extract_median_2IV(df)
  n_iv1 <- nlevels(df$condition1) 
  n_iv2 <- nlevels(df$condition2) 
  n_conditions <- n_iv1*n_iv2
  data_median <- c()
  for(i in 1:n_conditions){
    data_median <- append(data_median, c(median_period[i], median_period[i]))
  }
  # Create correctly named factors (here: prime and mask) and specify the same levels and labels as above
  data_medians <- tibble(period= data_median,
                         survival = rep(c(.5, 0),n_conditions),
                         prime =  factor(rep(rep(1:n_iv2, each=2),n_iv1),levels=c(1:3),labels=c("blank","congruent","incongruent")),
                         mask = factor(rep(1:n_iv1, each=2*n_iv2),levels=c(1:4),labels=c("nomask", "relevant", "irrelevant","lines")))

# plot the hazard functions
p1 <-df %>% 
  mutate(mask = condition1,       # Create correctly named factor
         prime = condition2) %>%  # Create correctly named factor
  ggplot(aes(x=period, color=prime, group=prime)) + # Apply correctly named factor
  geom_line(aes(y=hazard)) +
  geom_point(aes(y=hazard), size=1) + labs(color="Prime") +
  geom_linerange(aes(ymin=hazard-se_haz, ymax=hazard+se_haz), show.legend = F) +
  scale_x_continuous(breaks = c(1:(cutoff/binsize)), labels=c(1:(cutoff/binsize)*binsize),
                     limits = c(1,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="", y="h(t)", title = paste("Subject ", subj)) +
  theme(legend.background = element_rect(fill = "transparent"),
        panel.grid = element_blank(),
        legend.position = "top",
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~mask,nrow=1,ncol=4) # Apply correctly named factor

# plot the survivor functions
p2 <-df %>%
  mutate(mask = condition1,      # Create correctly named factor
         prime = condition2) %>% # Create correctly named factor
  ggplot(aes(x=period, color=prime, group=prime)) + # Apply correctly named factor
  geom_line(aes(y=survival), show.legend = F) +
  geom_point(aes(y=survival), size=1, show.legend = F) +
  geom_linerange(aes(ymin=survival-se_surv, ymax=survival+se_surv), show.legend = F) +
  # add vertical lines to indicate the estimated median RTs with geom_path()
  geom_path(aes(x=period, y=survival),
            data = data_medians, 
            linetype = 3, show.legend = F) +
  scale_x_continuous(breaks = c(1:(cutoff/binsize)), labels=c(1:(cutoff/binsize)*binsize),
                     limits=c(1,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="", y="S(t)") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~mask,nrow=1,ncol=4) # Apply correctly named factor

# plot the conditional accuracy functions
p3 <-df %>%
  mutate(mask = condition1,       # Create correctly named factor
         prime = condition2) %>%  # Create correctly named factor
  ggplot(aes(x=period, color=prime, group=prime)) + # Apply correctly named factor
  geom_line(aes(y=ca), show.legend = F) +
  geom_point(aes(y=ca), size=1, show.legend = F) +
  geom_linerange(aes(ymin=ca-se_ca, ymax=ca+se_ca), show.legend = F) +
  scale_x_continuous(breaks = c(1:(cutoff/binsize)), labels=c(1:(cutoff/binsize)*binsize),
                     limits=c(1,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="Time bin t's endpoint (ms)", y="ca(t)") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~mask,nrow=1,ncol=4) # Apply correctly named factor

p1/p2/p3
}
```

@@

```{r save-plots, warning=F}
# save the plot for each subject
map2(paste0("Tutorial_4_descriptive_2IV/Plot_2IV_for_subject", data_final$pid, "_PanisSchmidt.png"), data_final$plot, ggsave, width = 12, height = 8, dpi = 600)
```

>>>>

---
title: "Tutorial_1b"
author: "Sven Panis"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

@@

Tutorial_1b illustrates how to use functional programming to set up life tables, and plot the discrete-time hazard, survivor, and conditional accuracy functions, for the data of Experiment 1 of Panis and Schmidt (2016) when there are TWO independent variables (here: mask type and prime type).

@@

```{r read-raw-data}
raw <- read_delim(file="Tutorial_1_descriptive_stats/data/DataExp1_6subjects.txt", delim="\t",col_names=T) # 9372 rows
head(raw)
```

@@

# Apply functional programming to set up life tables, and plot the discrete-time functions (h(t), S(t), P(t), and ca(t)).

@@

```{r load-functions}
source("custom_functions.R")
```

@@

```{r save-plots, warning=F, eval=F}
# save the plot for each subject
map2(paste0("Tutorial_1_descriptive_stats/figures/Plot_2IV_for_subject", data_final$pid, "_PanisSchmidt.png"), data_final$plot, ggsave, width = 12, height = 10, dpi = 300)
```
